---
title: "Within--Between Decomposition for Repeated-Measure Algorithm Validation"
author: "Junrui"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Goal

Suppose we have repeated measurements for two variables:

- $Y_{ij}$: ground truth measurement for subject $i$ at time $j$
- $X_{ij}$: algorithm prediction of the ground truth for subject $i$ at time $j$

The scientific question is whether the algorithm prediction $X$ tracks the ground truth $Y$ over time.

A key complication is that both $X$ and $Y$ may vary across subjects and within subjects over time. A single pooled association between $X$ and $Y$ can therefore mix two different sources of information:

1. **Between-subject association**: Do subjects with higher average algorithm predictions also have higher average ground truth values?
2. **Within-subject association**: When a subject's algorithm prediction is higher than their own usual level, is their ground truth also higher?

This distinction is important for avoiding Simpson's paradox, ecological fallacy, or symptom-paradox-type interpretation problems.

# Mathematical setup

For each subject $i$, define the subject-level mean of $X$ as:

$$
\bar X_i = \frac{1}{n_i}\sum_{j=1}^{n_i} X_{ij}
$$

and the within-subject deviation as:

$$
X_{ij} - \bar X_i
$$

Then $X_{ij}$ can be decomposed as:

$$
X_{ij} = \bar X_i + (X_{ij} - \bar X_i)
$$

A within--between mixed model can be written as:

$$
Y_{ij}
= \beta_0
+ \beta_B \bar X_i
+ \beta_W (X_{ij} - \bar X_i)
+ \gamma t_{ij}
+ b_i
+ \epsilon_{ij}
$$

where:

- $\beta_B$ is the **between-subject slope**
- $\beta_W$ is the **within-subject slope**
- $t_{ij}$ is time
- $b_i$ is a subject-level random intercept
- $\epsilon_{ij}$ is residual error

The between-subject slope answers:

> Do people with higher average algorithm predictions also have higher average ground truth values?

The within-subject slope answers:

> When the same person's algorithm prediction is above their own average, is their ground truth also above their own average?

For algorithm validation, $\beta_W$ is often particularly important because it reflects within-person tracking over time.

# Why standardization is not required for testing $\beta_B = \beta_W$

Although $\bar X_i$ and $X_{ij} - \bar X_i$ often have different variances, they have the same measurement unit.

For example, suppose $X$ is an algorithm-predicted symptom score measured in points. Then:

- $X_{ij}$ is measured in symptom-score points
- $\bar X_i$ is measured in symptom-score points
- $X_{ij} - \bar X_i$ is also measured in symptom-score points

Therefore:

$$
\beta_B = \frac{\text{units of }Y}{\text{one point of }X}
$$

and:

$$
\beta_W = \frac{\text{units of }Y}{\text{one point of }X}
$$

So the contrast:

$$
H_0: \beta_B - \beta_W = 0
$$

is meaningful on the original scale.

Standardizing $\bar X_i$ and $X_{ij} - \bar X_i$ separately would answer a different question:

> Is a one-standard-deviation difference in between-subject $X$ associated with the same change in $Y$ as a one-standard-deviation within-subject deviation in $X$?

That can be useful descriptively, but it is not the same as testing whether a single raw-unit pooled slope is justified.

# Pooled model and equality of slopes

The pooled model is:

$$
Y_{ij}
= \beta_0
+ \beta X_{ij}
+ \gamma t_{ij}
+ b_i
+ \epsilon_{ij}
$$

This model implicitly assumes that the between-subject and within-subject slopes are the same:

$$
\beta_B = \beta_W = \beta
$$

So before using a pooled model, it is useful to test:

$$
H_0: \beta_B = \beta_W
$$

or equivalently:

$$
H_0: \beta_B - \beta_W = 0
$$

# Reparameterized model

An algebraically equivalent model is:

$$
Y_{ij}
= \beta_0
+ \beta_W X_{ij}
+ \delta \bar X_i
+ \gamma t_{ij}
+ b_i
+ \epsilon_{ij}
$$

where:

$$
\delta = \beta_B - \beta_W
$$

Therefore, testing whether the coefficient of $\bar X_i$ is zero in this reparameterized model is equivalent to testing:

$$
H_0: \beta_B = \beta_W
$$

This is usually the cleanest model-based test.

# Simulation setup

We simulate three scenarios:

1. **Opposite direction**: $\beta_B > 0$ and $\beta_W < 0$  
   This is the textbook Simpson's paradox / symptom paradox case.

2. **Same direction, different size**: $\beta_B > 0$ and $\beta_W > 0$, but $\beta_B \ne \beta_W$  
   This means the algorithm is directionally consistent, but between-subject and within-subject associations differ materially.

3. **Same direction, similar size**: $\beta_B \approx \beta_W > 0$  
   This is the case where a pooled model may be more defensible.

# Load packages

```{r packages}
library(lme4)
library(lmerTest)
library(multcomp)
library(ggplot2)
library(dplyr)
library(broom.mixed)
```

# Simulation function

```{r simulation-function}
simulate_within_between <- function(
    n_id = 120,
    n_time = 6,
    beta_B = 1.0,
    beta_W = 1.0,
    beta_time = 0.2,
    sigma_subject = 1.0,
    sigma_error = 1.0,
    mean_x_sd = 2.0,
    within_x_sd = 1.0,
    scenario_name = "scenario"
) {
  
  dat <- expand.grid(
    id = 1:n_id,
    time = 0:(n_time - 1)
  )
  
  dat <- dat %>%
    arrange(id, time)
  
  # Subject-level average X: between-subject component
  x_bar_true <- rnorm(n_id, mean = 0, sd = mean_x_sd)
  
  # Subject-level random intercept for Y
  subject_re <- rnorm(n_id, mean = 0, sd = sigma_subject)
  
  dat$x_bar_true <- x_bar_true[dat$id]
  dat$subject_re <- subject_re[dat$id]
  
  # Within-subject deviation of X
  dat$x_within_true <- rnorm(nrow(dat), mean = 0, sd = within_x_sd)
  
  # Observed X
  dat$x <- dat$x_bar_true + dat$x_within_true
  
  # Generate Y using separate between and within slopes
  dat$y <- 
    10 +
    beta_B * dat$x_bar_true +
    beta_W * dat$x_within_true +
    beta_time * dat$time +
    dat$subject_re +
    rnorm(nrow(dat), mean = 0, sd = sigma_error)
  
  # In real data, we usually do not know the true x_bar or true within component.
  # Therefore, we estimate them empirically from observed X.
  dat <- dat %>%
    group_by(id) %>%
    mutate(
      x_bar = mean(x),
      x_within = x - x_bar
    ) %>%
    ungroup()
  
  dat$scenario <- scenario_name
  
  return(dat)
}
```

# Model fitting function

```{r fit-models-function}
fit_models <- function(dat) {
  
  # Model A: Explicit within-between decomposition
  # y = beta_B * x_bar + beta_W * x_within + time + random intercept
  m_wb <- lmer(
    y ~ x_bar + x_within + time + (1 | id),
    data = dat
  )
  
  # Model B: Reparameterized version
  # y = beta_W * x + delta * x_bar + time + random intercept
  # where delta = beta_B - beta_W
  m_reparam <- lmer(
    y ~ x + x_bar + time + (1 | id),
    data = dat
  )
  
  # Model C: Pooled model
  # Assumes beta_B = beta_W
  m_pooled <- lmer(
    y ~ x + time + (1 | id),
    data = dat
  )
  
  list(
    within_between = m_wb,
    reparameterized = m_reparam,
    pooled = m_pooled
  )
}
```

# Summary function

```{r summary-function}
summarize_results <- function(models, scenario_label) {
  
  m_wb <- models$within_between
  m_reparam <- models$reparameterized
  m_pooled <- models$pooled
  
  wb_coef <- summary(m_wb)$coefficients
  rep_coef <- summary(m_reparam)$coefficients
  pooled_coef <- summary(m_pooled)$coefficients
  
  beta_B_hat <- wb_coef["x_bar", "Estimate"]
  beta_W_hat <- wb_coef["x_within", "Estimate"]
  
  beta_B_p <- wb_coef["x_bar", "Pr(>|t|)"]
  beta_W_p <- wb_coef["x_within", "Pr(>|t|)"]
  
  # In the reparameterized model:
  # coefficient of x is beta_W
  # coefficient of x_bar is beta_B - beta_W
  beta_W_reparam <- rep_coef["x", "Estimate"]
  delta_hat <- rep_coef["x_bar", "Estimate"]
  delta_p <- rep_coef["x_bar", "Pr(>|t|)"]
  
  pooled_beta <- pooled_coef["x", "Estimate"]
  pooled_p <- pooled_coef["x", "Pr(>|t|)"]
  
  out <- data.frame(
    scenario = scenario_label,
    
    beta_B_hat_decomposed = beta_B_hat,
    beta_W_hat_decomposed = beta_W_hat,
    beta_B_minus_beta_W_from_decomposed = beta_B_hat - beta_W_hat,
    
    beta_W_hat_reparameterized = beta_W_reparam,
    delta_beta_B_minus_beta_W = delta_hat,
    p_value_test_beta_B_equals_beta_W = delta_p,
    
    pooled_beta_hat = pooled_beta,
    pooled_beta_p_value = pooled_p
  )
  
  return(out)
}
```

# Simulate three scenarios

```{r simulate-scenarios}
set.seed(2026)

# Scenario 1:
# Simpson's paradox / symptom paradox
# Between effect is positive, within effect is negative.
dat1 <- simulate_within_between(
  beta_B = 1.2,
  beta_W = -0.8,
  beta_time = 0.2,
  scenario_name = "1_opposite_direction"
)

# Scenario 2:
# Same direction, but different magnitude.
dat2 <- simulate_within_between(
  beta_B = 1.2,
  beta_W = 0.3,
  beta_time = 0.2,
  scenario_name = "2_same_direction_different_size"
)

# Scenario 3:
# Same direction, similar magnitude.
dat3 <- simulate_within_between(
  beta_B = 0.8,
  beta_W = 0.75,
  beta_time = 0.2,
  scenario_name = "3_same_direction_similar_size"
)
```

# Fit models

```{r fit-scenarios}
models1 <- fit_models(dat1)
models2 <- fit_models(dat2)
models3 <- fit_models(dat3)
```

# Summarize main results

```{r main-results}
results <- bind_rows(
  summarize_results(models1, "1_opposite_direction"),
  summarize_results(models2, "2_same_direction_different_size"),
  summarize_results(models3, "3_same_direction_similar_size")
)

print(results)
```

# Interpretation of model coefficients

For the decomposed model:

```r
y ~ x_bar + x_within + time + (1 | id)
```

- `x_bar` estimates $\beta_B$
- `x_within` estimates $\beta_W$

For the reparameterized model:

```r
y ~ x + x_bar + time + (1 | id)
```

- `x` estimates $\beta_W$
- `x_bar` estimates $\beta_B - \beta_W$

For the pooled model:

```r
y ~ x + time + (1 | id)
```

- `x` estimates a single pooled slope, implicitly assuming $\beta_B = \beta_W$

# Inspect model summaries

```{r model-summaries}
summary(models1$within_between)
summary(models1$reparameterized)
summary(models1$pooled)

summary(models2$within_between)
summary(models2$reparameterized)
summary(models2$pooled)

summary(models3$within_between)
summary(models3$reparameterized)
summary(models3$pooled)
```

# Contrast test using `glht()`

The decomposed model estimates $\beta_B$ and $\beta_W$ separately. We can use `multcomp::glht()` to test:

$$
H_0: \beta_B - \beta_W = 0
$$

In the fitted model, this is:

```r
x_bar - x_within = 0
```

```{r glht-function}
contrast_beta_B_minus_beta_W_glht <- function(model) {
  
  # This tests:
  # H0: beta_B - beta_W = 0
  #
  # In the decomposed model:
  # beta_B = coefficient of x_bar
  # beta_W = coefficient of x_within
  
  glht_result <- glht(
    model,
    linfct = c("x_bar - x_within = 0")
  )
  
  return(glht_result)
}
```

```{r glht-run}
glht1 <- contrast_beta_B_minus_beta_W_glht(models1$within_between)
glht2 <- contrast_beta_B_minus_beta_W_glht(models2$within_between)
glht3 <- contrast_beta_B_minus_beta_W_glht(models3$within_between)

summary(glht1)
confint(glht1)

summary(glht2)
confint(glht2)

summary(glht3)
confint(glht3)
```

# Clean table from `glht()` results

```{r glht-table-function}
extract_glht_contrast <- function(glht_object, scenario_label) {
  
  test_sum <- summary(glht_object)
  ci_sum <- confint(glht_object)
  
  out <- data.frame(
    scenario = scenario_label,
    contrast = names(test_sum$test$coefficients),
    estimate_beta_B_minus_beta_W = as.numeric(test_sum$test$coefficients),
    se = as.numeric(test_sum$test$sigma),
    z_value = as.numeric(test_sum$test$tstat),
    p_value = as.numeric(test_sum$test$pvalues),
    lower_ci = as.numeric(ci_sum$confint[, "lwr"]),
    upper_ci = as.numeric(ci_sum$confint[, "upr"])
  )
  
  return(out)
}
```

```{r glht-table}
glht_results <- bind_rows(
  extract_glht_contrast(glht1, "1_opposite_direction"),
  extract_glht_contrast(glht2, "2_same_direction_different_size"),
  extract_glht_contrast(glht3, "3_same_direction_similar_size")
)

print(glht_results)
```

# Compare `glht()` with reparameterized model

The `glht()` estimate from the decomposed model should match the `x_bar` coefficient from the reparameterized model:

```{r compare-glht-reparam}
reparam_results <- bind_rows(
  tidy(models1$reparameterized, effects = "fixed") %>%
    filter(term == "x_bar") %>%
    mutate(scenario = "1_opposite_direction"),
  tidy(models2$reparameterized, effects = "fixed") %>%
    filter(term == "x_bar") %>%
    mutate(scenario = "2_same_direction_different_size"),
  tidy(models3$reparameterized, effects = "fixed") %>%
    filter(term == "x_bar") %>%
    mutate(scenario = "3_same_direction_similar_size")
) %>%
  select(scenario, term, estimate, std.error, statistic, p.value)

comparison_glht_reparam <- glht_results %>%
  select(
    scenario,
    glht_estimate = estimate_beta_B_minus_beta_W,
    glht_p_value = p_value
  ) %>%
  left_join(
    reparam_results %>%
      select(
        scenario,
        reparam_estimate = estimate,
        reparam_p_value = p.value
      ),
    by = "scenario"
  )

print(comparison_glht_reparam)
```

# Clean pooled-model comparison table

```{r clean-results}
clean_results <- results %>%
  mutate(
    pooling_decision = case_when(
      abs(delta_beta_B_minus_beta_W) < 0.15 ~ "Pooling likely reasonable",
      TRUE ~ "Do not pool; slopes differ materially"
    )
  )
print(clean_results)
```

The threshold of `0.15` above is only an illustrative rule. In a real validation analysis, the threshold for a practically meaningful difference should be defined based on the measurement scale, clinical context, and intended use of the algorithm.

# Plot raw repeated-measure relationship

This plot shows the raw relationship between $X$ and $Y$.

- Thin lines connect repeated measurements within subject.
- The thick smooth line shows the pooled trend.

```{r raw-plot-function}
plot_raw_relationship <- function(dat, title_text) {
  dat %>% mutate(id = as.character(id)) %>%
  ggplot(., aes(x = x, y = y, group = id, color = id)) +
    geom_line(alpha = 0.25) +
    geom_point(alpha = 0.45) +
    geom_smooth(
      aes(group = 1),
      method = "lm",
      se = FALSE,
      linewidth = 1.2
    ) +
    guides(color = "none") + 
    labs(
      title = title_text,
      subtitle = "Thin lines = within-subject repeated measures; thick line = pooled trend",
      x = "Algorithm prediction X",
      y = "Ground truth Y"
    ) +
    theme_bw()
}
```

```{r raw-plots}
plot_raw_relationship(dat1, "Scenario 1: Opposite between and within effects")
plot_raw_relationship(dat2, "Scenario 2: Same direction but different magnitude")
plot_raw_relationship(dat3, "Scenario 3: Same direction and similar magnitude")
```

# Plot within-subject relationship

This plot centers both $X$ and $Y$ within each subject.

```{r within-plot-function}
plot_within_relationship <- function(dat, title_text) {
  
  dat <- dat %>%
    group_by(id) %>%
    mutate(
      y_bar = mean(y),
      y_within = y - y_bar
    ) %>%
    ungroup()
  
  ggplot(dat, aes(x = x_within, y = y_within)) +
    geom_point(alpha = 0.45) +
    geom_smooth(
      method = "lm",
      se = FALSE,
      linewidth = 1.2
    ) +
    labs(
      title = title_text,
      subtitle = "Both X and Y are centered within subject",
      x = "Within-subject deviation in X",
      y = "Within-subject deviation in Y"
    ) +
    theme_bw()
}
```

```{r within-plots}
plot_within_relationship(dat1, "Scenario 1: Within-subject association")
plot_within_relationship(dat2, "Scenario 2: Within-subject association")
plot_within_relationship(dat3, "Scenario 3: Within-subject association")
```

# Plot between-subject relationship

This plot collapses each subject to one point.

```{r between-plot-function}
plot_between_relationship <- function(dat, title_text) {
  
  subject_level <- dat %>%
    group_by(id) %>%
    summarize(
      x_bar = mean(x),
      y_bar = mean(y),
      .groups = "drop"
    )
  
  ggplot(subject_level, aes(x = x_bar, y = y_bar)) +
    geom_point(alpha = 0.65) +
    geom_smooth(
      method = "lm",
      se = FALSE,
      linewidth = 1.2
    ) +
    labs(
      title = title_text,
      subtitle = "Each point is one subject's average X and average Y",
      x = "Subject mean algorithm prediction",
      y = "Subject mean ground truth"
    ) +
    theme_bw()
}
```

```{r between-plots}
plot_between_relationship(dat1, "Scenario 1: Between-subject association")
plot_between_relationship(dat2, "Scenario 2: Between-subject association")
plot_between_relationship(dat3, "Scenario 3: Between-subject association")
```

# Optional: compare unstandardized and standardized effects

This section illustrates why separate standardization changes the question.

```{r standardized-effects-function}
calculate_standardized_effects <- function(dat, models, scenario_label) {
  
  m_wb <- models$within_between
  beta_B <- fixef(m_wb)["x_bar"]
  beta_W <- fixef(m_wb)["x_within"]
  
  sd_x_bar <- sd(unique(dat[, c("id", "x_bar")])$x_bar)
  sd_x_within <- sd(dat$x_within)
  
  data.frame(
    scenario = scenario_label,
    beta_B_raw = beta_B,
    beta_W_raw = beta_W,
    sd_x_bar = sd_x_bar,
    sd_x_within = sd_x_within,
    beta_B_per_1SD_x_bar = beta_B * sd_x_bar,
    beta_W_per_1SD_x_within = beta_W * sd_x_within
  )
}
```

```{r standardized-effects}
standardized_effects <- bind_rows(
  calculate_standardized_effects(dat1, models1, "1_opposite_direction"),
  calculate_standardized_effects(dat2, models2, "2_same_direction_different_size"),
  calculate_standardized_effects(dat3, models3, "3_same_direction_similar_size")
)

print(standardized_effects)
```

The raw slopes answer:

> Is one raw unit of $X$ associated with the same change in $Y$ between subjects and within subjects?

The standardized effects answer:

> Is one standard deviation of between-subject $X$ associated with the same change in $Y$ as one standard deviation of within-subject $X$?

These are related but different questions.

# Scenario-specific interpretation

## Scenario 1: Opposite direction

True data-generating values:

$$
\beta_B = 1.2, \quad \beta_W = -0.8
$$

This is the most dangerous case. The algorithm may separate high-burden from low-burden subjects, but it may fail to track within-person change correctly, or even track it in the wrong direction.

In this scenario, the pooled model is misleading because it compresses two opposing associations into one coefficient.

## Scenario 2: Same direction, different size

True data-generating values:

$$
\beta_B = 1.2, \quad \beta_W = 0.3
$$

Both associations are positive, but the between-subject association is much stronger than the within-subject association.

This means the algorithm may rank subjects well, but it may be less sensitive to within-person change.

The pooled model may be directionally correct, but it hides an important difference between cross-sectional validity and longitudinal tracking validity.

## Scenario 3: Same direction, similar size

True data-generating values:

$$
\beta_B = 0.8, \quad \beta_W = 0.75
$$

Here the between-subject and within-subject slopes are similar. If the estimated difference is small and the confidence interval excludes practically meaningful differences, the pooled model may be defensible.

# Practical recommendation

For repeated-measure algorithm validation, a useful workflow is:

1. Fit the within--between model:

$$
Y_{ij}
= \beta_0
+ \beta_B \bar X_i
+ \beta_W (X_{ij} - \bar X_i)
+ \gamma t_{ij}
+ b_i
+ \epsilon_{ij}
$$

2. Report $\hat\beta_B$, $\hat\beta_W$, and $\hat\beta_B - \hat\beta_W$.

3. Use either the reparameterized model or `glht()` to test:

$$
H_0: \beta_B - \beta_W = 0
$$

4. Fit the pooled model only if the between and within slopes are sufficiently similar.

5. Do not rely only on a non-significant p-value. Consider the confidence interval and whether the possible difference is practically meaningful for the intended algorithm use.

